First steps in synthetic algebraic geometry

### 2017-07-21 slides-duesseldorf2017.tex
Slides of today's talk

\title{First steps in synthetic algebraic geometry}
\author{Ingo Blechschmidt}
\date{February 7th, 2017}

@@

  Algebraic Geometry Seminar \\
  University of Warwick \\
  \ \\
  Feburary 7th, 2017
  \par
\end{frame}

\frame[t]{\frametitle{Outline}\small
  \begin{columns}[t]
    \begin{column}{0.6\textwidth}
      \vspace*{-1.5em}
      \begin{itemize}\item[]\tableofcontents\end{itemize}
      \par
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering
      \begin{minipage}{0.70\textwidth}
        \begin{exampleblock}{}
          \justifying
          Any finitely generated vector space does \emph{not not} possess a basis.
        \end{exampleblock}
      \end{minipage}
      \medskip

      \begin{minipage}{0.70\textwidth}
        \begin{exampleblock}{}
          \justifying
          Any sheaf of modules of finite type on a reduced scheme is locally free
          \emph{on a dense open subset}.
        \end{exampleblock}
        \centering
        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
        \par
      \end{minipage}
    \end{column}
  \end{columns}
}

\note{\justifying\fontsize{8pt}{9.6}\selectfont
  \begin{center}\large\textbf{Abstract}\end{center}

  \begin{changemargin}{2.5em}{2.5em}
    (new abstract)
  \end{changemargin}
}

@@

  Further topics regarding the little Zariski topos:
  \begin{itemize}
    \item Transfer principles~$M \leftrightarrow M^\sim$
    \item Understanding generic freeness
    \item The curious role of affine open subsets
    \item Quasicoherence
    \item Spreading from points to neighbourhoods
    \item The relative spectrum
  \end{itemize}
\end{frame}

@@

\subsection[Generic freeness]{Generic freeness}

\begin{frame}\frametitle{\mbox{A curious property of the structure sheaf}}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

  \centering
  \includegraphics[scale=0.3]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
  \tiny
  %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
  %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
  %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
  %pp.~189--210.
  Miles Tierney. On the spectrum of a ringed topos. 1976.
  \par
\end{frame}

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there
  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  is $f \neq 0$ in $R$ such that
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
  \begin{itemize}
    \item $\O_{\Spec R}$ is a \hil{field}: $\neg(\text{$x$ invertible}) \Rightarrow x = 0$,
    \item $\O_{\Spec R}$ and $\O_{\Spec R}[X_1,\ldots,X_n]$ are \hil{Noetherian}
    in the sense that any ideal is \hil{not not} finitely generated.
  \end{itemize}
\end{frame}

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip
  \pause

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \small
  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

@@

  \begin{itemize}
    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
    \item $\Delta \defeq \RelSpec \affl[\varepsilon]/(\varepsilon^2) \cong \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$
  \end{itemize}\bigskip

@@

  \item An $\affl$-module~$E$ is quasicoherent if and only if
  \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
  is an isomorphism for all finitely presented $\affl$-algebras $R$.

@@

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \in \affl$ such that
  $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.

@@

  \item If $\gamma : \Delta \to X$ is tangent vector with $\gamma(0) \in U$,
  then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.
  \bigskip\pause

>>>>

\title{First steps in synthetic algebraic geometry}
\author{Ingo Blechschmidt}
\date{July 21st, 2017}

@@

  Oberseminar Algebra/Geometrie/Topologie \\
  Heinrich-Heine-Universität Düsseldorf \\
  \ \\
  July 21st, 2017
  \par
\end{frame}

\backupstart
{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/unterbacher-see}}
\begin{frame}[plain]
  \vspace{\textheight}\vspace{-0em}
  \centering
  \textcolor{white}{\emph{Camping at Unterbacher See}} \\
  \par
\end{frame}}

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Reduced rings are Noetherian and in fact fields.
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
  \end{enumerate}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

%\frame[t]{\frametitle{Outline}\small
%  \begin{columns}[t]
%    \begin{column}{0.6\textwidth}
%      \vspace*{-1.5em}
%      \begin{itemize}\item[]\tableofcontents\end{itemize}
%      \par
%    \end{column}
%    \begin{column}{0.4\textwidth}
%      \centering
%      \begin{minipage}{0.70\textwidth}
%        \begin{exampleblock}{}
%          \justifying
%          Any finitely generated vector space does \emph{not not} possess a basis.
%        \end{exampleblock}
%      \end{minipage}
%      \medskip
%
%      \scalebox{3}{$\Downarrow$}
%
%      \begin{minipage}{0.70\textwidth}
%        \begin{exampleblock}{}
%          \justifying
%          Any sheaf of modules of finite type on a reduced scheme is locally free
%          \emph{on a dense open subset}.
%        \end{exampleblock}
%        \centering
%        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
%        \par
%      \end{minipage}
%    \end{column}
%  \end{columns}
%}
%
%\note{\justifying\fontsize{8pt}{9.6}\selectfont
%  \begin{center}\large\textbf{Abstract}\end{center}
%
%  \begin{changemargin}{2.5em}{2.5em}
%    (new abstract)
%  \end{changemargin}
%}

@@

  Further topics regarding the little Zariski topos:
  \begin{itemize}
    \item Unique features of the internal world
    \item Transfer principles~$M \leftrightarrow M^\sim$
    \item The curious role of affine open subsets
    \item Quasicoherence
    \item Spreading from points to neighbourhoods
    \item The relative spectrum
  \end{itemize}
\end{frame}

\subsection[Unique features]{Unique features of the internal world}

\begin{frame}\frametitle{Unique features of the internal world}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

  \only<1>{
    \centering
    \includegraphics[scale=0.3]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
    \tiny
    %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
    %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
    %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
    %pp.~189--210.
    Miles Tierney. On the spectrum of a ringed topos. 1976.
    \par
  }
  \pause

  If~$X$ is reduced, this implies:
  \begin{itemize}
    \item $\O_X$ is a \hil{field} in that $\neg(\text{$x$ invertible})
    \Rightarrow x = 0$.
    \item $\O_X$ has \hil{$\neg\neg$-stable equality}: $\neg\neg(x = 0)
    \Rightarrow x = 0$.
    \item $\O_X$ is \hil{anonymously Noetherian}: Any ideal of~$\O_X$ is \mbox{\hil{not
    not}} finitely generated.
  \end{itemize}
\end{frame}

\subsection[Generic freeness]{Grothendieck's generic freeness lemma}

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  \hil{Lemma.} If $R$ is reduced, there
  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  is $f \neq 0$ in $R$ such that
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
  $\O_{\Spec R}$ is a \hil{field} and \hil{anonymously Noetherian}.
\end{frame}

@@

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \? A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip
  \pause

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

@@

  \begin{itemize}
    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
    \item $\Delta \defeq \RelSpec \affl[\varepsilon]/(\varepsilon^2) \cong \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$
  \end{itemize}\bigskip

@@

  \item An $\affl$-module~$E$ is \hil{quasicoherent} if and only if
  \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
  is an isomorphism for all finitely presented $\affl$-algebras $R$.

@@

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \? \affl$ such that
  $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.

@@

  \item If $\gamma : \Delta \to X$ is a tangent vector with $\gamma(0) \in U$,
  then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.
  \bigskip\pause
